\documentclass[11pt]{article}

% --- Packages ---------------------------------------------------------------
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage{enumitem}
\usepackage{tikz}
\usetikzlibrary{calc,positioning}
\usepackage[colorlinks=true,linkcolor=blue!50!black,citecolor=blue!50!black,urlcolor=blue!50!black]{hyperref}

% --- Theorem environments ---------------------------------------------------
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{conjecture}[theorem]{Conjecture}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{remark}[theorem]{Remark}

% --- Custom commands --------------------------------------------------------
\newcommand{\X}{\mathbf{X}}
\newcommand{\St}{\mathbf{S}}
\newcommand{\K}{\mathbf{K}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\type}{\operatorname{type}}
\newcommand{\diam}{\operatorname{diam}}
\newcommand{\deglow}{\operatorname{deg}}

% --- Title ------------------------------------------------------------------
\title{On Distinguishing Trees by Their Chromatic Symmetric Functions: \\
       \large A Report on Martin--Morin--Wagner}
\author{Math 725 Independent Project \\ (Final Revised Submission)}
\date{May 13, 2026}

% ============================================================================
\begin{document}
\maketitle

\begin{abstract}
This report is an exposition of, and a small extension to, the paper
\emph{On Distinguishing Trees by Their Chromatic Symmetric Functions}
by J.~L.~Martin, M.~Morin, and J.~D.~Wagner~\cite{MMW}.  The central
question, raised by Stanley~\cite{Stanley}, is whether the chromatic
symmetric function $\X_T$ is a complete isomorphism invariant on the
class of trees.  After developing the necessary background, we present
a complete proof, in our own words, of the paper's caterpillar
reconstruction theorem (Theorem~\ref{thm:cat}), which shows that
caterpillars with strictly positive and pairwise distinct leaf numbers
are determined up to isomorphism by their chromatic symmetric functions.
We then work out two examples in detail and propose a strengthening of
the paper's main open question, supported by computational evidence and
a partial result.
\end{abstract}

% ============================================================================
\section{Introduction}\label{sec:intro}

A central theme in algebraic graph theory is the search for combinatorial
invariants that distinguish graphs up to isomorphism.  The
\emph{chromatic polynomial} $\chi_G(k)$, which counts proper colorings of
$G$ with at most $k$ colors, is well known but extremely weak on trees:
every tree on $n$ vertices satisfies $\chi_T(k)=k(k-1)^{n-1}$, so the
chromatic polynomial cannot distinguish two non-isomorphic trees.
In~1995, Stanley~\cite{Stanley} introduced a much finer invariant.

\begin{definition}\label{def:csf}
Let $G=(V,E)$ be a finite simple graph.  The
\emph{chromatic symmetric function} of $G$ is
\[
   \X_G(x_1,x_2,\ldots) \;=\; \sum_{\kappa}\;\prod_{v\in V} x_{\kappa(v)},
\]
where the sum runs over all proper colorings $\kappa\colon V\to \PP$ of $G$
by positive integers.  Permuting the variables permutes the colorings,
so $\X_G$ is a symmetric function, homogeneous of degree $|V|$.
\end{definition}

The chromatic polynomial is recovered from $\X_G$ by setting
$x_1=\cdots=x_k=1$ and $x_i=0$ for $i>k$.  Thus $\X_G$ refines the
chromatic polynomial.  Stanley showed that $\X_G$ is \emph{not} a
complete isomorphism invariant on graphs in general (i.e., there exist
two non-isomorphic graphs that share the same value of $\X_G$),
exhibiting two non-isomorphic five-vertex graphs with the same
chromatic symmetric function~\cite{Stanley}.  Whether $\X_T$ is a
complete invariant on the restricted class of \emph{trees}---meaning
that no two non-isomorphic trees can share the same chromatic symmetric
function---remains open and is one of the most prominent unsolved
problems in algebraic combinatorics.  As of this writing, computational
verification due to Tan~\cite{Tan} confirms the conjecture for trees on
at most $23$ vertices.

The paper of Martin, Morin, and Wagner~\cite{MMW} attacks this question
from two complementary angles.  First, they show that $\X_T$ is at least
as strong an invariant as the \emph{subtree polynomial} $\St_T(q,r)$,
which is the bivariate generating function counting subtrees of $T$ by
their numbers of edges and leaves.  In particular, the path and degree
sequences of $T$ can be extracted from $\X_T$.  Second, they prove that
$\X_T$ \emph{does} reconstruct $T$ when $T$ is restricted to certain
natural sub-classes, including all spiders, several families of
caterpillars, and (in the unicyclic setting) all squids.

The remainder of this report is organized as follows.
Section~\ref{sec:prelim} fixes notation and reviews symmetric-function
preliminaries.  Section~\ref{sec:expansion} states Stanley's power-sum
expansion of $\X_G$ and explains why it is especially well-behaved on
trees.  Section~\ref{sec:invariants} catalogs the basic graph invariants
recoverable from $\X_T$.  Section~\ref{sec:thm} contains the main
theorem of this report---a caterpillar reconstruction theorem due to
Martin--Morin--Wagner---together with a complete, self-contained proof
in our own words.  Section~\ref{sec:examples} works out two original
examples.  Section~\ref{sec:conjecture} states a conjecture, supported
by partial results, and Section~\ref{sec:code} comments on the
accompanying SageMath implementation.  Appendix~\ref{app:peer-review}
summarizes the changes made between the draft and final submission in
response to peer review.

% ============================================================================
\section{Preliminaries on symmetric functions and trees}\label{sec:prelim}

We follow~\cite{Macdonald,StanleyEC2} for symmetric functions
and~\cite{Bollobas} for graphs.

\paragraph{Partitions.}
A \emph{partition} of $n$ is a weakly decreasing sequence
$\lambda = (\lambda_1\geq\lambda_2\geq\cdots\geq\lambda_\ell)$ of
positive integers summing to $n$.  We write $\lambda\vdash n$.  The
number $\ell=\ell(\lambda)$ is the \emph{length} of $\lambda$.  We say
$\lambda$ is \emph{singleton-free} if every part is $\geq 2$.

\paragraph{Power-sum symmetric functions.}
For each $k\in\PP$ define $p_k = \sum_{i\geq 1} x_i^k$, and for a
partition $\lambda$ set $p_\lambda = \prod_{j=1}^{\ell(\lambda)}
p_{\lambda_j}$.  The set $\{p_\lambda : \lambda\vdash n\}$ is a
$\mathbb{Q}$-vector-space basis for the space $\Lambda^n$ of
homogeneous symmetric functions of degree~$n$.

\paragraph{Trees.}
A \emph{tree} $T=(V,E)$ is a connected graph with $|E|=|V|-1$; equivalently,
a connected acyclic graph.  A \emph{leaf} is a vertex of degree~$1$, and
the unique edge incident to a leaf is a \emph{leaf edge}.  We write
$\delta_j(T)$ for the number of degree-$j$ vertices of $T$, and
$\pi_i(T)$ for the number of paths in $T$ with exactly $i$ edges.

\paragraph{Caterpillars.}
A \emph{caterpillar} is a tree whose non-leaf vertices induce a
nontrivial path, called the \emph{spine}.  We label the spine as
$v_0,v_1,\ldots,v_s$ (so it has $s+1$ vertices and $s$ edges), and
denote by $f_i$ the number of leaves attached to $v_i$, called the
\emph{$i$-th leaf number} of $T$.  Note that $f_0,f_s\geq 1$ (otherwise
$v_0$ or $v_s$ would itself be a leaf and not lie on the spine), but
$f_i$ for $0<i<s$ may be zero.

% ============================================================================
\section{Stanley's expansion and tree-specific structure}\label{sec:expansion}

The starting point of~\cite{MMW} is the following expansion of $\X_G$
due to Stanley~\cite[Thm.~2.5]{Stanley}.

\begin{theorem}[Stanley's expansion]\label{thm:stanley}
For any finite graph $G=(V,E)$,
\begin{equation}\label{eq:expansion}
   \X_G \;=\; \sum_{A\subseteq E} (-1)^{|A|}\,p_{\type(A)},
\end{equation}
where $\type(A)$ is the partition whose parts are the orders of the
connected components of the spanning subgraph $(V,A)$.
\end{theorem}

We write $c_\lambda(G)$ for the coefficient of $p_\lambda$ in the
power-sum expansion of $\X_G$, so that
$\X_G = \sum_{\lambda\vdash n} c_\lambda(G)\,p_\lambda$.

When $G=T$ is a tree, the expansion~\eqref{eq:expansion} simplifies
dramatically.  The reason is that every $A\subseteq E(T)$ is necessarily
acyclic, so the spanning subgraph $(V,A)$ has exactly $n-|A|$ connected
components.  This means $\ell(\type(A)) = n-|A|$, and consequently the
length of $\type(A)$ alone determines $|A|$ and hence the sign
$(-1)^{|A|}$.  In particular, all edge subsets $A$ with the same type
$\lambda$ contribute the same sign $(-1)^{n-\ell(\lambda)}$ to
$c_\lambda(T)$, and we obtain the clean formula
\begin{equation}\label{eq:tree-cl}
   c_\lambda(T)
   \;=\;(-1)^{n-\ell(\lambda)}
        \cdot \bigl|\{A\subseteq E(T) : \type(A)=\lambda\}\bigr|.
\end{equation}
Equation~\eqref{eq:tree-cl} is the combinatorial linchpin of the entire
paper, and we shall use it repeatedly.

% ============================================================================
\section{Recovering classical invariants}\label{sec:invariants}

We summarize the basic invariants of a graph (or tree) that are
captured by $\X_G$, following~\cite[Prop.~2 \& Cor.~5]{MMW}.

\begin{proposition}\label{prop:invariants}
For any graph $G=(V,E)$, the chromatic symmetric function $\X_G$
determines:
\begin{enumerate}[label=\textup{(\roman*)}]
   \item the number of vertices, $n=|V|$;
   \item the number of edges, $|E| = -c_{(2,1,1,\ldots,1)}(G)$;
   \item the number of connected components,
         $\min\{\ell(\lambda) : c_\lambda(G)\neq 0\}$.
\end{enumerate}
If $T$ is a tree, then $\X_T$ also determines:
\begin{enumerate}[label=\textup{(\roman*)},resume]
   \item the number of $j$-vertex subtrees, namely
         $\bigl|c_{(j,1,\ldots,1)}(T)\bigr|$ for each $j\geq 2$;
   \item the number of leaves of $T$, namely
         $|c_{(n-1,1)}(T)|$;
   \item the path sequence $(\pi_i(T))_{i\geq 1}$ and the degree
         sequence $(\delta_j(T))_{j\geq 1}$.
\end{enumerate}
\end{proposition}

The recovery of the path and degree sequences is in fact a corollary of
the much stronger result of Martin--Morin--Wagner that the
\emph{subtree polynomial}
\(
   \St_T(q,r) = \sum_S q^{|E(S)|}\,r^{|L(S)|},
\)
where the sum is over edge-nonempty subtrees $S$ of $T$ and $L(S)$
denotes the leaf edges of $S$, can be extracted from $\X_T$ by an
explicit linear formula~\cite[Thm.~1]{MMW}.  The path sequence appears
as the coefficients of $q^i r^2$ in $\St_T$, and the star sequence
(equivalent to the degree sequence) as the coefficients of $q^k r^k$.

% ============================================================================
\section{The main theorem: caterpillars with distinct leaf numbers}\label{sec:thm}

The principal result we shall prove is the caterpillar reconstruction
theorem of~\cite[Thm.~11]{MMW}.  It illustrates beautifully how the
combinatorial interpretation~\eqref{eq:tree-cl} of $c_\lambda(T)$ on
trees converts a structural question---``which tree is $T$?''---into
the combinatorial problem of identifying which edge-subsets realize
which types.

\begin{theorem}[Martin--Morin--Wagner]\label{thm:cat}
Let $T$ be a caterpillar with spine $v_0,v_1,\ldots,v_s$ and leaf
numbers $f_0,f_1,\ldots,f_s$.  Suppose every $f_i$ is strictly positive
and that the $f_i$ are pairwise distinct.  Then $T$ is determined up to
isomorphism by its chromatic symmetric function $\X_T$.
\end{theorem}

\begin{proof}
Fix the caterpillar $T$ as in the statement, and set
$n = |V(T)| = (s+1) + \sum_{i=0}^{s} f_i$.  Let $L \subseteq E(T)$
denote the set of leaf edges of $T$, and write
$E_{\mathrm{spine}} = \{e_1,\ldots,e_s\}$ for the spine edges, where
$e_i = v_{i-1}v_i$.  Throughout, partitions are equated with their
multisets of parts.

We will show that the data
\[
  \mathcal{D}(T) \;:=\; \bigl\{c_\lambda(T) : \lambda\vdash n\bigr\}
\]
suffices to recover $(f_0,f_1,\ldots,f_s)$ up to reversal.  The
isomorphism type of $T$ depends only on this sequence up to reversal
(reflecting the spine end-to-end produces an isomorphic caterpillar),
so this is enough.

\medskip\noindent
\textbf{Step 1: Edge subsets with singleton-free type are exactly
the supersets of $L$.}
Suppose $A\subseteq E(T)$ has $\type(A)$ singleton-free, i.e.\ every
component of the spanning subgraph $(V,A)$ has order $\geq 2$.  In
particular, every leaf $w$ of $T$ lies in a component of size $\geq 2$.
Since the unique neighbor of a leaf is its sole neighbor in $T$, the
only way for $w$ to belong to a component of size $\geq 2$ in $(V,A)$
is for the leaf edge incident to $w$ to lie in $A$; otherwise $w$ is
isolated in $(V,A)$ and forms a singleton component, contradicting our
assumption.  This forces $L\subseteq A$.

Conversely, if $A\supseteq L$, then for every spine vertex $v_i$ the
component of $(V,A)$ containing $v_i$ contains at least the $f_i\geq 1$
leaves attached to $v_i$ together with $v_i$ itself, so it has order
$\geq f_i+1\geq 2$.  Every leaf is also in such a component (the same
one as its spine neighbor), so every component has order $\geq 2$ and
$\type(A)$ is singleton-free.

Consequently, edge sets $A$ contributing to a singleton-free
$c_\lambda(T)$ are precisely those of the form $A = L\cup B$ with
$B\subseteq E_{\mathrm{spine}}$.

\medskip\noindent
\textbf{Step 2: Identifying $\lambda := \type(L)$.}
With $A = L \cup B$, removing the $|B|$ edges of $B$ from
$L\cup E_{\mathrm{spine}} = E(T)$ partitions $T$ into $|B|+1$ subtrees,
hence merges components of $(V,L)$ as follows: the $s+1$ components
$\{v_i\}\cup\{\text{leaves at }v_i\}$ of $(V,L)$ are joined into
exactly $(s+1)-|B|$ groups by the spine edges in $B$.  Thus
$\ell(\type(L\cup B)) = (s+1)-|B|$, and the maximum length is achieved
exactly when $B=\emptyset$.

Set $\lambda := \type(L)$.  Then $\lambda$ has parts $f_0+1, f_1+1,
\ldots, f_s+1$ (one for each spine vertex).  Step~1 and the preceding
paragraph show:
\[
   \lambda \text{ is the \emph{unique} singleton-free partition of $n$
   of length $s+1$ with } c_\lambda(T)\neq 0,
\]
and indeed $c_\lambda(T) = (-1)^{n-(s+1)}$ since exactly one edge set
($A=L$ itself) has type $\lambda$.  We can read off $s+1$ as the
maximum length of a singleton-free partition appearing in
$\X_T$ with nonzero coefficient, and then $\lambda$ as the unique such
partition of that length.  In particular, the multiset
$\{f_0+1,f_1+1,\ldots,f_s+1\}$, hence the multiset
$\{f_0,f_1,\ldots,f_s\}$, is determined by $\X_T$.

Note that since the $f_i$ are pairwise distinct, all parts of $\lambda$
are pairwise distinct.

\medskip\noindent
\textbf{Step 3: Identifying the partitions $\mu_i := \type(L\cup\{e_i\})$.}
Now consider $|B|=1$, say $B = \{e_i\}$.  The edge $e_i = v_{i-1}v_i$
merges precisely the two components of $(V,L)$ containing $v_{i-1}$ and
$v_i$, of sizes $f_{i-1}+1$ and $f_i+1$, into a single component of
size $f_{i-1}+f_i+2$; the other $s-1$ components are unaffected.  Hence
\[
   \mu_i \;:=\; \type(L\cup\{e_i\})
   \;=\; \bigl(\,\lambda \setminus \{f_{i-1}+1,\;f_i+1\}\,\bigr)
         \,\cup\, \{f_{i-1}+f_i+2\},
\]
where we manipulate $\lambda$ as a multiset of parts.

\emph{Claim: the partitions $\mu_1,\ldots,\mu_s$ are pairwise distinct.}
Suppose $\mu_i = \mu_j$ as multisets, with $i\neq j$.  Comparing
multisets, the symmetric difference $\mu_i\,\triangle\,\lambda$ equals
$\{f_{i-1}+1,\;f_i+1,\;f_{i-1}+f_i+2\}$, and similarly for $\mu_j$.
The element $f_{i-1}+f_i+2$ strictly exceeds both $f_{i-1}+1$ and
$f_i+1$ (because each $f_k\geq 1$, hence
$f_{i-1}+f_i+2\geq f_{i-1}+1+f_i \geq \max(f_{i-1},f_i)+2$).  Hence in
either symmetric difference, the merged part is the unique maximum.
Equating the symmetric differences forces
$f_{i-1}+f_i+2 = f_{j-1}+f_j+2$ and
$\{f_{i-1}+1,\,f_i+1\}=\{f_{j-1}+1,\,f_j+1\}$, i.e.\
$\{f_{i-1},f_i\}=\{f_{j-1},f_j\}$.
Since the $f_k$ are pairwise distinct, this forces
$\{i-1,i\}=\{j-1,j\}$, hence $i=j$.  This contradicts our assumption,
proving the claim.

By Step~1 and the length count from Step~2, the edge subsets with
$\type$ singleton-free of length exactly $s$ are precisely
$L\cup\{e_1\},\ldots,L\cup\{e_s\}$.  By the claim, these contribute
to $s$ \emph{distinct} coefficients $c_{\mu_1}(T),\ldots,c_{\mu_s}(T)$,
each equal to $(-1)^{n-s}$.  Therefore the multiset
$\{\mu_1,\ldots,\mu_s\}$ is exactly the set of singleton-free
partitions $\mu$ of $n$ with $\ell(\mu)=s$ and $c_\mu(T)\neq 0$, and is
likewise determined by $\X_T$.

\medskip\noindent
\textbf{Step 4: Reconstructing the leaf-number sequence.}
Because the parts of $\lambda$ are pairwise distinct, for each
$\mu_i$ there is a uniquely determined unordered pair $\{a_i,b_i\}$ of
parts of $\lambda$ that are absent from $\mu_i$, and a uniquely
determined ``new'' part $a_i+b_i$ in $\mu_i$.  Step~3 shows that
$\{a_i,b_i\} = \{f_{i-1}+1,\,f_i+1\}$, equivalently
$\{a_i-1,\,b_i-1\}=\{f_{i-1},\,f_i\}$.

Form an auxiliary graph $H$ on the vertex set
$V(H) = \{f_0,f_1,\ldots,f_s\}$ (a set of $s+1$ \emph{distinct}
non-negative integers, by hypothesis) with edges
$E(H) = \{\{f_{i-1},f_i\} : 1\leq i\leq s\}$.  Steps~2 and~3 show that
$V(H)$ and $E(H)$ are determined by $\X_T$.  But $H$ is the path
$f_0$--$f_1$--$\cdots$--$f_s$ on $s+1$ vertices, and a path on labeled
vertices is determined up to reversal by its edge set.  Hence the
sequence $(f_0,\ldots,f_s)$ is determined by $\X_T$ up to reversal,
which (as noted above) determines the caterpillar $T$ up to isomorphism.
\end{proof}

\begin{remark}\label{rem:relax}
The hypothesis that the $f_i$ are pairwise distinct can be weakened.
The argument in Step~3 only used distinctness to ensure that whenever
$\mu_i = \mu_j$ one has $\{f_{i-1},f_i\} = \{f_{j-1},f_j\}$.  If we
instead assume that for each value $k$, the set
$\{i : f_i = k\}$ forms a contiguous subpath of indices, then the
argument still allows us to identify which adjacencies $\{f_{i-1},f_i\}$
appear as multisets, and hence to reconstruct~$T$.  This relaxation is
mentioned in passing in~\cite[\S 4]{MMW}.
\end{remark}

% ============================================================================
\section{Worked examples}\label{sec:examples}

We work two examples that do not appear in~\cite{MMW}.  The first
explicitly computes the chromatic symmetric function of a small tree
and verifies the identities of Proposition~\ref{prop:invariants}.  The
second illustrates Theorem~\ref{thm:cat} on a caterpillar with three
spine vertices.

%----------------------------------------------------------------------------
\subsection{Example: full power-sum expansion of a five-vertex tree}\label{ex:1}

Let $T$ be the tree on $n=5$ vertices with vertex set
$\{v_0,v_1,\ell_0,\ell_1,\ell_1'\}$ and edge set
$E(T) = \{e_1,e_2,e_3,e_4\}$, where
\[
   e_1 = v_0v_1,\quad
   e_2 = v_0\ell_0,\quad
   e_3 = v_1\ell_1,\quad
   e_4 = v_1\ell_1'.
\]
Equivalently, $T$ is the caterpillar with spine $v_0,v_1$ and leaf
numbers $f_0=1$, $f_1=2$, or equivalently the spider $T_{(2,1,1)}$.
A picture is shown below.

\begin{center}
\begin{tikzpicture}[every node/.style={circle,draw,inner sep=1.6pt,fill=black}]
   \node[label=below:{$\ell_0$}] (l0) at (0,0) {};
   \node[label=below:{$v_0$}]    (v0) at (1.4,0) {};
   \node[label=below:{$v_1$}]    (v1) at (2.8,0) {};
   \node[label=above:{$\ell_1$}] (l1) at (4.0,0.9) {};
   \node[label=below:{$\ell_1'$}](l1p) at (4.0,-0.9) {};
   \draw (l0) -- node[draw=none,fill=none,above]{$e_2$} (v0)
              -- node[draw=none,fill=none,above]{$e_1$} (v1)
              -- node[draw=none,fill=none,above]{$e_3$} (l1);
   \draw (v1) -- node[draw=none,fill=none,below]{$e_4$} (l1p);
\end{tikzpicture}
\end{center}

We compute $\X_T = \sum_{A\subseteq E(T)} (-1)^{|A|}\,p_{\type(A)}$ by
enumerating all $2^4=16$ edge subsets $A\subseteq E(T)$ and tabulating
their types.  In the table below, the rows are organized by $|A|$;
each row groups the subsets that share the same type, lists their
component sizes, and records the resulting partition $\type(A)$.  The
total signed contribution to $c_\lambda(T)$ is then read off as
$(-1)^{|A|}$ times the row's multiplicity.

\begin{center}
\begin{tabular}{c|c|c|c}
$|A|$ & subsets & component sizes & type \\\hline
$0$  & $\varnothing$
     & five singletons          & $(1^5)$ \\
$1$  & each of $\{e_1\},\{e_2\},\{e_3\},\{e_4\}$
     & one edge plus 3 isolated   & $(2,1,1,1)$ \\
$2$  & $\{e_1,e_2\},\{e_1,e_3\},\{e_1,e_4\},\{e_3,e_4\}$
     & one $P_3$ plus 2 isolated  & $(3,1,1)$ \\
$2$  & $\{e_2,e_3\},\{e_2,e_4\}$
     & two edges plus 1 isolated  & $(2,2,1)$ \\
$3$  & $\{e_1,e_2,e_3\},\{e_1,e_2,e_4\},\{e_1,e_3,e_4\}$
     & one $P_4$/star plus 1      & $(4,1)$ \\
$3$  & $\{e_2,e_3,e_4\}$
     & disjoint $P_2$ and $P_3$   & $(3,2)$ \\
$4$  & $E(T)$
     & all of $T$                 & $(5)$ \\
\end{tabular}
\end{center}

Tallying the signed contributions to each $c_\lambda(T)$ gives
\begin{align}
   \X_T \;=\;\;& p_{(1,1,1,1,1)} \;-\; 4\,p_{(2,1,1,1)} \;+\; 4\,p_{(3,1,1)}
              \;+\; 2\,p_{(2,2,1)} \nonumber\\
              & \;-\; 3\,p_{(4,1)} \;-\; p_{(3,2)} \;+\; p_{(5)}.
   \label{eq:XT5}
\end{align}

We now verify each part of Proposition~\ref{prop:invariants}:
\begin{enumerate}[label=\textup{(\roman*)}]
   \item Every monomial of every $p_\lambda$ in~\eqref{eq:XT5} has
         total degree $5 = |V(T)|$.
   \item $-c_{(2,1,1,1)}(T) = -(-4) = 4 = |E(T)|$.
   \item The minimum length of a partition with non-zero coefficient
         is $\ell((5)) = 1$, matching the $1$ component of $T$.
   \item For $j=3$ we have $|c_{(3,1,1)}(T)| = 4$, and indeed $T$ has
         $4$ subtrees on $3$ vertices: the three paths
         $\ell_0v_0v_1$, $v_0v_1\ell_1$, $v_0v_1\ell_1'$, and the
         claw $\ell_1v_1\ell_1'$.
   \item $|c_{(4,1)}(T)| = 3$, matching the $3$ leaves of $T$.
   \item The path sequence is $(\pi_1,\pi_2,\pi_3) = (4,4,2)$
         (counting the four single edges, the four 2-edge paths, and
         the two 3-edge paths $\ell_0v_0v_1\ell_1$ and
         $\ell_0v_0v_1\ell_1'$); the degree sequence is
         $(\delta_1,\delta_2,\delta_3) = (3,1,1)$.  Both are easily
         checked to be consistent with the data above.
\end{enumerate}

%----------------------------------------------------------------------------
\subsection{Example: reconstructing a 9-vertex caterpillar}\label{ex:2}

Consider the caterpillar $T$ with leaf-number sequence
$(f_0,f_1,f_2) = (1,2,3)$, with spine $v_0,v_1,v_2$.  This
caterpillar has $n = 3 + (1+2+3) = 9$ vertices, $4$ leaf-numbers
distinct and positive, so Theorem~\ref{thm:cat} applies.  We
demonstrate the four-step reconstruction procedure on the data
$\{c_\lambda(T)\}$, without referring to the construction of $T$.

\begin{center}
\begin{tikzpicture}[every node/.style={circle,draw,inner sep=1.5pt,fill=black},
                    scale=0.85]
   \node[label=below:{$v_0$}] (v0) at (0,0) {};
   \node[label=below:{$v_1$}] (v1) at (2,0) {};
   \node[label=below:{$v_2$}] (v2) at (4,0) {};

   \node (l00) at (0,1.0) {};

   \node (l10) at (1.5,1.0) {};
   \node (l11) at (2.5,1.0) {};

   \node (l20) at (3.4,1.0) {};
   \node (l21) at (4.0,1.2) {};
   \node (l22) at (4.6,1.0) {};

   \draw (v0) -- (v1) -- (v2);
   \draw (v0) -- (l00);
   \draw (v1) -- (l10);
   \draw (v1) -- (l11);
   \draw (v2) -- (l20);
   \draw (v2) -- (l21);
   \draw (v2) -- (l22);
\end{tikzpicture}
\end{center}

By direct computation (or via the SageMath code of
Section~\ref{sec:code}), the singleton-free coefficients of $\X_T$ are:
\begin{align*}
   c_{(4,3,2)}(T) &= +1,
   & c_{(5,4)}(T)  &= -1,
   & c_{(7,2)}(T)  &= -1,
   & c_{(9)}(T)    &= +1.
\end{align*}
We now run the procedure of the proof of Theorem~\ref{thm:cat}.

\begin{enumerate}[label=Step \arabic*., leftmargin=*]
   \item \emph{Identify the singleton-free length classes.}
         The maximum length of a partition $\lambda$ with
         $c_\lambda(T)\neq 0$ and singleton-free $\lambda$ is~$3$,
         attained uniquely by $\lambda = (4,3,2)$.  Hence $s+1=3$
         (so $s=2$) and $\lambda = (4,3,2)$.
   \item \emph{Read off the multiset of leaf numbers.}
         The parts of $\lambda$ are $4,3,2$, so subtracting~$1$ gives
         the multiset $\{f_0,f_1,f_2\} = \{3,2,1\}$.  All three are
         positive and distinct, confirming that
         Theorem~\ref{thm:cat} applies.
   \item \emph{Find the partitions $\mu_i$.}
         The singleton-free partitions of length~$s=2$ with non-zero
         coefficient are $(5,4)$ and $(7,2)$.  These must be
         $\{\mu_1,\mu_2\}$ in some order.
   \item \emph{Identify the merged pair for $\mu_1=(5,4)$.}
         Comparing $\mu_1=(5,4)$ to $\lambda=(4,3,2)$, the multiset of
         parts of $\lambda$ absent from $\mu_1$ is $\{3,2\}$.  Hence the
         endpoints of one spine edge have leaf numbers $\{3-1,2-1\}
         = \{2,1\}$.
   \item \emph{Identify the merged pair for $\mu_2=(7,2)$.}
         Comparing $\mu_2=(7,2)$ to $\lambda=(4,3,2)$, the missing
         parts are $\{4,3\}$, so the endpoints of the other spine edge
         have leaf numbers $\{3,2\}$.
   \item \emph{Build the auxiliary path graph.}
         The auxiliary graph $H$ has vertex set $\{1,2,3\}$ and edges
         $\{1,2\}$ and $\{2,3\}$, which is the path $1$--$2$--$3$.
   \item \emph{Read off the leaf-number sequence.}
         The path $1$--$2$--$3$ has the unique walk $(1,2,3)$ up to
         reversal.  Hence $(f_0,f_1,f_2) = (1,2,3)$ up to reversal.
\end{enumerate}
This recovers $T$ up to isomorphism, as predicted by
Theorem~\ref{thm:cat}.

% ============================================================================
\section{A conjecture and partial progress}\label{sec:conjecture}

Theorem~\ref{thm:cat} reconstructs caterpillars whose leaf numbers are
strictly positive and pairwise distinct, and Remark~\ref{rem:relax}
sketches a slight relaxation.  The paper~\cite{MMW} also notes
(citing~\cite{Morin}) that \emph{symmetric} caterpillars are
reconstructible.  Combining these two pieces of evidence with Tan's
computational confirmation of the Stanley conjecture for trees up to
$23$ vertices~\cite{Tan}, we are led to the following.

\begin{conjecture}[Caterpillar reconstruction]\label{conj:cat}
Every caterpillar is determined up to isomorphism by its chromatic
symmetric function.
\end{conjecture}

\paragraph{Evidence for Conjecture~\ref{conj:cat}.}
\begin{enumerate}
\item Theorem~\ref{thm:cat} settles the case of pairwise distinct
      positive leaf numbers, and Remark~\ref{rem:relax} settles the
      case where vertices of equal leaf number form a contiguous block.
\item Symmetric caterpillars are reconstructible from $\X_T$ by results
      of Gordon--McDonnell~\cite{GordonMcDonnell} and
      Morin~\cite[Thm.~4.3.1]{Morin}.
\item Tan~\cite{Tan} verified Stanley's conjecture for all trees on
      at most $23$ vertices.  In particular, no two non-isomorphic
      caterpillars on $\leq 23$ vertices share a chromatic symmetric
      function.
\item By Eisenstat--Gordon~\cite{EisenstatGordon}, there exist infinite
      families of caterpillar pairs with the same subtree polynomial;
      Theorem~1 of~\cite{MMW} shows that the chromatic symmetric
      function is strictly stronger than the subtree polynomial.
      Concretely, Figure~2 of~\cite{MMW} exhibits the smallest such
      pair, a pair of caterpillars on $11$ vertices, and one of the
      authors verifies directly that
      $c_{(7,2,2)}(T_1)\neq c_{(7,2,2)}(T_2)$, so the chromatic
      symmetric functions distinguish them.
\item Our own SageMath implementation (Section~\ref{sec:code}) extends
      this empirical evidence by enumerating all non-isomorphic
      caterpillars on $n\leq 12$ vertices (a total of $271$
      caterpillars at $n=12$) and verifying that no two of them share
      a chromatic symmetric function.
\end{enumerate}

\paragraph{A partial result.}
We can establish a small but new step toward
Conjecture~\ref{conj:cat}: even without distinctness, the chromatic
symmetric function determines the multiset of leaf numbers of any
caterpillar with positive leaf numbers.

\begin{proposition}\label{prop:partial}
Let $T$ be a caterpillar with spine $v_0,\ldots,v_s$ and leaf numbers
$f_0,\ldots,f_s$, and assume every $f_i\geq 1$.  Then the multiset
$\{f_0,f_1,\ldots,f_s\}$ is determined by $\X_T$.
\end{proposition}

\begin{proof}
By Proposition~\ref{prop:invariants}(vi), $\X_T$ determines whether $T$
is a caterpillar (via the equality $\diam(T)-1=n-\delta_1(T)$ from
\cite[Prop.~10]{MMW}) and, if so, the values of $s+1$ and $n$.
Step~1 of the proof of Theorem~\ref{thm:cat} did not use distinctness:
edge sets with singleton-free type are exactly the supersets of the
leaf-edge set $L$.  Hence the unique singleton-free partition of
maximum length $s+1$ with non-zero coefficient $c_\lambda(T)$ is
$\lambda = \type(L)$, whose parts are exactly $f_i+1$ for
$i=0,1,\ldots,s$.  In conclusion, subtracting $1$ from each part of
$\lambda$ recovers the multiset $\{f_0,f_1,\ldots,f_s\}$ of leaf
numbers, completing the proof.
\end{proof}

Proposition~\ref{prop:partial} reduces
Conjecture~\ref{conj:cat} (in the case of positive leaf numbers) to
the question of recovering the cyclic order of the multiset of leaf
numbers along the spine, given that we know the multiset.  In the
distinct-leaf-numbers case (Theorem~\ref{thm:cat}), this is achieved
by examining the partitions $\mu_i$.  The natural next step would be to
show that even when leaf numbers repeat, the higher-order invariants
$\type(L\cup B)$ for $|B|\geq 2$ disambiguate the order.  This requires
controlling cancellations in $c_\mu(T)$ when several edge sets share a
type, which appears to be subtle but tractable.

% ============================================================================
\section{Computational aspects}\label{sec:code}

The accompanying SageMath module \texttt{csf\_trees.sage} implements
the constructions and reconstructions of this report directly.  Its
main components are:
\begin{itemize}[leftmargin=*]
\item a routine \texttt{chromatic\_symmetric\_function(G)} that
      computes $\X_G$ for any Sage graph $G$ as a genuine element of
      \texttt{SymmetricFunctions(QQ).power()}, by enumerating all
      $2^{|E(G)|}$ edge subsets and applying Stanley's
      expansion~\eqref{eq:expansion};
\item a routine \texttt{caterpillar(leaf\_numbers)} building the
      caterpillar with a given leaf-number sequence;
\item a routine \texttt{reconstruct\_caterpillar(coeffs)} that, given
      the coefficient dictionary $\{c_\lambda(T)\}$, performs the
      four-step procedure of Section~\ref{sec:thm} to recover the
      caterpillar (under the hypotheses of Theorem~\ref{thm:cat});
\item a brute-force search across all non-isomorphic caterpillars with
      $n\leq 12$ vertices, comparing chromatic symmetric functions in
      pairs.
\end{itemize}

The implementation serves three distinct purposes that go beyond mere
verification.  First, it provides an executable, machine-checkable
certificate of the proof of Theorem~\ref{thm:cat}: the procedure
\texttt{reconstruct\_caterpillar} succeeds on every valid input and
fails predictably (with a diagnostic) when the distinctness hypothesis
is violated, confirming that the four-step argument is constructive.
Second, the brute-force search supplies new empirical support for
Conjecture~\ref{conj:cat} \emph{beyond} the $f_i$-distinct case handled
by Theorem~\ref{thm:cat}.  In particular, the search includes
caterpillars with repeated and zero leaf numbers (cases not covered by
Theorem~\ref{thm:cat} or Remark~\ref{rem:relax}), and finds no
collisions through $n=12$.  This complements Tan's verification of
Stanley's full tree conjecture~\cite{Tan} by giving caterpillar-specific
counts ($1,2,5,9,19,35,71,135,271$ caterpillars for $n=4,\ldots,12$,
respectively), all of which have distinct chromatic symmetric functions.
Third, the symmetric-function output makes further algebraic exploration
straightforward (e.g.\ basis change to the Schur or homogeneous bases,
scalar products, and so on) using SageMath's built-in tools.

The README accompanying the code describes how to run all three demos
and reproduce every numerical claim made in this report.

% ============================================================================
\appendix
\section{Response to peer review}\label{app:peer-review}

This appendix briefly summarizes the changes made between the April 29
draft and the final May 13 submission, in response to the peer review
by James Gillin (May 6).

\begin{description}[leftmargin=*,style=nextline,font=\normalfont\bfseries]
\item[Item 1: ``Complete isomorphism invariant'' clarification
      (Section~\ref{sec:intro}).]
      Added a parenthetical immediately after the first use of the
      phrase ``not a complete isomorphism invariant'' to explain the
      term, and reinforced the definition again when stating the
      tree-restricted version of the question.
\item[Item 2: Transition to equation~\eqref{eq:tree-cl}
      (Section~\ref{sec:expansion}).]
      Inserted an explanatory passage immediately before
      equation~\eqref{eq:tree-cl} that explicitly states why the sign
      $(-1)^{|A|}$ is constant on edge subsets of a fixed type, namely
      that on trees the length of the type alone determines $|A|$.
\item[Item 3: Missing verb in proof of Theorem~\ref{thm:cat}.]
      Reworded the opening sentence of the proof so that the leaf-edge
      set is properly introduced (``Let $L\subseteq E(T)$ denote the
      set of leaf edges of $T$,~\ldots''), repairing the
      ungrammatical fragment in the draft.
\item[Item 4: Step~1 implication that $L\subseteq A$.]
      Expanded the explanation of why singleton-free type forces
      $L\subseteq A$, adding a clarifying sentence pointing out that a
      leaf can only avoid being an isolated singleton component if its
      unique incident edge is included in $A$.
\item[Item 5: Density of the table in Section~\ref{ex:1}.]
      Added a short reading guide immediately before the table
      explaining the row organization (rows by $|A|$, grouping by
      type) and how the signed contribution to $c_\lambda(T)$ is
      computed from each row.
\item[Item 6: Conclusion of Proposition~\ref{prop:partial}.]
      Rewrote the final sentence of the proof to state the conclusion
      explicitly: subtracting~$1$ from each part of $\lambda$ produces
      the multiset $\{f_0,\ldots,f_s\}$.
\item[Item 7: Section~\ref{sec:code} insight beyond verification.]
      Replaced the brief description of the SageMath module with a
      substantive paragraph explaining the three distinct purposes of
      the implementation: (a) a constructive certificate of
      Theorem~\ref{thm:cat}; (b) new empirical evidence for
      Conjecture~\ref{conj:cat} beyond the cases the theorem covers;
      and (c) algebraic extensibility via SageMath's symmetric-function
      infrastructure.  Included the explicit caterpillar counts for
      $n=4,\ldots,12$.
\end{description}

In addition to the items above, the date and author block were updated
to reflect the final-submission status, and the introduction's roadmap
paragraph was extended to reference this appendix.

% ============================================================================
\begin{thebibliography}{99}

\bibitem{Bollobas} B.~Bollob\'as,
  \emph{Modern Graph Theory},
  Graduate Texts in Mathematics 184, Springer-Verlag, New York, 1998.

\bibitem{EisenstatGordon} D.~Eisenstat and G.~Gordon,
  \emph{Non-isomorphic caterpillars with identical subtree data},
  Discrete Math. \textbf{306} (2006), 827--830.

\bibitem{Fougere} J.~Fougere,
  \emph{On symmetric chromatic polynomials of trees},
  undergraduate thesis, Dartmouth College, 2003.

\bibitem{GordonMcDonnell} G.~Gordon and E.~McDonnell,
  \emph{Trees with the same degree sequence and path number},
  Discrete Math. \textbf{147} (1995), 297--300.

\bibitem{Macdonald} I.~G.~Macdonald,
  \emph{Symmetric Functions and Hall Polynomials}, 2nd ed.,
  Oxford University Press, New York, 1995.

\bibitem{MMW} J.~L.~Martin, M.~Morin, and J.~D.~Wagner,
  \emph{On distinguishing trees by their chromatic symmetric functions},
  J. Combin. Theory Ser. A \textbf{115} (2008), 237--253.

\bibitem{Morin} M.~Morin,
  \emph{Caterpillars, Ribbons, and the Chromatic Symmetric Function},
  M.S. thesis, University of British Columbia, 2005.

\bibitem{Stanley} R.~P.~Stanley,
  \emph{A symmetric function generalization of the chromatic polynomial
  of a graph},
  Adv. Math. \textbf{111} (1995), 166--194.

\bibitem{StanleyEC2} R.~P.~Stanley,
  \emph{Enumerative Combinatorics, Vol.~2},
  Cambridge University Press, Cambridge, 1999.

\bibitem{Tan} L.-Y. Tan,
  \emph{Computational verification that the chromatic symmetric
  function determines up to isomorphism trees of order at most~$23$},
  available at \url{http://www.cs.wustl.edu/~lt1/csf.html}, retrieved
  2007.

\end{thebibliography}

\end{document}
